One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction

### 2017-12-04 draft_05.tex
working on the breathing mode section

Breathing frequency

@@

The gas contains $N$ such identical particles, each of mass $m$, in a harmonic trap.  The
Hamiltonian describing the gas is 
\begin{equation}
H=\frac{1}{2}m\omega_0^2\sum_{i}x_i^2+\frac{1}{2}m\sum_i
v_i^2+\sum_{\left|x_i-x_j\right|\textless\sigma}F_0\left(\sigma-\left|x_i-x_j\right|\right)
\label{eq:preparation2}
\end{equation}
%
It is useful to rescale the quantities, and measure distance, time, energy and force in units of
$\sigma$, $1/\omega$, $m\omega_0^2\sigma^2$ and $m\omega_0^2\sigma$ respectively:
\begin{equation} \label{eq:transform}
  \tilde{x_i} = \frac{x_i}{\sigma}\ , \quad 
  \tilde{H} = \frac{H}{m\omega_0^2\sigma^2}\ ,  \quad \tilde{t} = \omega_0t \ ,  \quad 
\tilde{F_0} = \frac{F_0}{m\omega_0^2\sigma}\ .
\end{equation}
%
Eq.~\ref{eq:preparation2} is then rewritten as
\begin{equation}
\tilde{H}=\frac{1}{2}\sum_{i}\tilde{x_i}^2+\frac{1}{2}\sum_i\left(\frac{d\tilde{x}}{d\tilde{t}}\right)^2+\sum_{\left|\tilde{x_i}-\tilde{x_j}\right|\textless 1}F_0\left(1-\left|\tilde{x_i}-\tilde{x_j}\right|\right)
\end{equation}
%
Through this rescaling, we have reduced the number of the parameters in our model to three: the
energy $\tilde{H}$, the interaction strength $\tilde{F_0}$, and the number of particles $N$.  The
rescaling is equivalent to setting $\sigma$, $\omega_0$ and $m$ to $1$; this is what we do in our
numerical simulation.  In the rest of this paper, we will use ``E" and ``$F_0$" to denote the
reduced versions $\tilde{H}$ and $\tilde{F_0}$.

@@

In the parameter regime where the energy $E$ is very large ($E\gg F_0$), it is dominated by the trap
energy, i.e., the interaction energy is negligible.  Thus,
$E\approx\omega_0^2\sum_{i}x_i^2=N\omega_0^2R^2$ in this regime

\begin{figure}[tbhp]
\centering
\includegraphics[scale=0.6]{ZhiyuPictures/N=5_GS_pre_2_rev.eps}
\caption{The ground state of a system with $N=5$ particles, shown via the positions of the
  particles.  At large interactions, the particles position themselves just outside the range of
  interactions of the neighboring particles.}
\label{fig:GS1}
\end{figure}

@@

The lowest-energy state of the system has zero kinetic energy; in this state the particles find
stationary positions which minimize the trap (potential) and interaction energies.  The trap
potential tries to squeeze the particles at the trap center, while the interaction tries to push
them apart.

Because of the `Heaviside theta' form of our interaction, at large enough interactions (large $F_0$)
the particles are spaced exactly at distance equal to the range $\sigma$, which is distance $1$ in
our rescaled units.  The interaction is effectively a `hard-core' interaction.

@@

In the Fig.\ref{fig:GS1}, we can see this crossover from the ``solid-like" limit (left) to the
``hardcore gas" limit (right).

@@

\begin{figure*}[h]
%\subfigure[$t=0$]{
%\begin{minipage}[b]{0.18\linewidth}
%\includegraphics[scale=0.21]{ZhiyuPictures/stationary_frame_t=1-5T_rev.eps} 
%\includegraphics[scale=0.22]{ZhiyuPictures/rotating_frame_t=1-5T_rev.eps}
%\end{minipage}
%}
%\subfigure[$t=\frac{1}{5}T$]{
%\begin{minipage}[b]{0.18\linewidth}
%\includegraphics[scale=0.21]{ZhiyuPictures/%stationary_frame_t=2-5T_rev.eps} 
%\includegraphics[scale=0.22]{ZhiyuPictures/rotating_frame_t=2-5T_rev.eps}
%\end{minipage}
%}\subfigure[$t=\frac{2}{5}T$]{
%\begin{minipage}[b]{0.18\linewidth}
%\includegraphics[scale=0.21]{ZhiyuPictures/stationary_frame_t=3-5T_rev.eps} 
%\includegraphics[scale=0.22]{ZhiyuPictures/rotating_frame_t=3-5T_rev.eps}
%\end{minipage}
%}\subfigure[$t=\frac{3}{5}T$]{
%\begin{minipage}[b]{0.18\linewidth}
%\includegraphics[scale=0.21]{ZhiyuPictures/stationary_frame_t=4-5T_rev.eps} 
%\includegraphics[scale=0.22]{ZhiyuPictures/rotating_frame_t=4-5T_rev.eps}
%\end{minipage}
%}\subfigure[$t=\frac{4}{5}T$]{
%\begin{minipage}[b]{0.18\linewidth}
%\includegraphics[scale=0.21]{ZhiyuPictures/stationary_frame_t=5-5T_rev.eps} 
%\includegraphics[scale=0.22]{ZhiyuPictures/rotating_frame_t=5-5T_rev.eps}
%\end{minipage}
%}
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/stationary_and_rotating_frame_t=1-5_5T.eps}

\caption{N=50,F0=100,E=50000. Upper(lower) five pictures are snapshot of the configuration of the
  cloud in the stationary(rotating) frame at t=$\frac{1}{5}T$, $\frac{2}{5}T$, $\frac{3}{5}T$,
  $\frac{4}{5}T$, $T$ respectively.
WHAT IS T?
}
\label{fig:Breathingfrequency2_0}
\end{figure*}

\begin{figure*}
%\includegraphics[scale=0.23]{ZhiyuPictures/10_18rotating_frame_t=0T_rev.eps} 
%\includegraphics[scale=0.23]{ZhiyuPictures/10_18rotating_frame_t=20T_rev.eps}
%\includegraphics[scale=0.23]{ZhiyuPictures/10_18rotating_frame_t=40T_rev.eps} 
%\includegraphics[scale=0.23]{ZhiyuPictures/10_18rotating_frame_t=60T_rev.eps}
%\includegraphics[scale=0.23]{ZhiyuPictures/10_18rotating_frame_t=80T_rev.eps}
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.eps}  
\caption{the precession behavior of cloud viewed in the rotating frame in phase-space. $\frac{t}{T}=0,20,40,60,80$}
\label{fig:Breathingfrequency2_1}
\end{figure*}

A very useful way to visualize the state and evolution of the gas is to plot the position and
momentum of each particle, i.e., to plot the locations of the particles in the single-particle phase
space.  This will be useful for visuualizing both the breathing mode and relaxation.

Figures \ref{fig:Breathingfrequency2_0} and \ref{fig:Breathingfrequency2_1} show such phase space
snapshots.

@@

Since the number of particles we use in the Molecular Dynamics simulation is limited (5 to 50), the
random noise is significant. If we start with a state whose configuration in phase-space is an
ellipse slightly deviated from circle, the oscillating amplitude will be too small to be
distinguished from noise. Therefore, we start from an extreme case -- line distribution (random x,
zero p) and observe the oscillation of R(t).
\begin{figure}[h]
\center
\includegraphics[scale=0.32]{ZhiyuPictures/freq_scanF_differentE_log_2_with_oscillation_demo.png}
\caption{breathing mode frequency measured at different E and $F_0$. Insert: a demonstration of oscillation, $E_{trap}$ is the total potential energy of particles in the trap, which is proportional to $\sum_{is}{x_i^2}$, thus manifest the oscillation behavior of radius R(t) defined in Eq.\ref{eq:def_of_R}. The breathing frequency is measured by taking Fourier transformation of R(t).}
\label{fig:Breathingfrequency1}
\end{figure}

Without interaction, the breathing mode frequency is exactly 2. When there is interaction, the
frequency will drift to some other value near 2. We measure the radius of the cloud $R(t)$ and get
the frequency spectrum of its oscillation behavior by Fourier transform. Then we take the peak
frequency near 2 as the breathing mode frequency. The frequency measured in different E and $F_0$ is
shown in the Fig.\ref{fig:Breathingfrequency1}.

The mechanism that deviate breathing frequency from 2 is simple in real space. When two particle
bounce, they exchange momentum immediately. It can be interpret as particle A carries its momentum
$P_A$ and jump $\sigma$ to the right, while particle B carries its momentum $P_B$ and jump $\sigma$
to the left. In this manner, every collision will save a particle some time $\frac{\sigma}{v}$
. Since $v$ is proportional to R while the number of collision each particle experiences in one
period of harmonic oscillation can be estimated by N, we will get $\delta\sim
N^{\frac{3}{2}}E^{-\frac{1}{2}}$.

@@

This rotating frame picture is actually a classical version of the interaction picture, which turns out to symplifly our dynamic analysis. The stationary frame and rotating frame is shown in Fig.\ref{fig:Breathingfrequency2_1}.

>>>>

Breathing frequency, comparisong to quantum case ------

@@

The gas contains $N$ such identical particles, each of mass $m$, in a harmonic trap.  The
Hamiltonian describing the gas is 
\begin{equation}
H=\frac{1}{2}m\omega_0^2\sum_{i}x_i^2+\frac{1}{2}m\sum_i
v_i^2+\sum_{\left|x_i-x_j\right|\textless\sigma}F_0\left(\sigma-\left|x_i-x_j\right|\right)
\label{eq:preparation2}
\end{equation}
%
It is useful to rescale the quantities.  We will measure distance, time, energy and force in units
of $\sigma$, $1/\omega$, $m\omega_0^2\sigma^2$ and $m\omega_0^2\sigma$ respectively:
\begin{equation} \label{eq:transform}
  \tilde{x_i} = \frac{x_i}{\sigma}\ , \quad 
  \tilde{H} = \frac{H}{m\omega_0^2\sigma^2}\ ,  \quad \tilde{t} = \omega_0t \ ,  \quad 
\tilde{F_0} = \frac{F_0}{m\omega_0^2\sigma}\ .
\end{equation}
%
Eq.~\ref{eq:preparation2} is then rewritten as
\begin{equation}
\tilde{H}=\frac{1}{2}\sum_{i}\tilde{x_i}^2+\frac{1}{2}\sum_i\left(\frac{d\tilde{x}_i}{d\tilde{t}}\right)^2+\sum_{\left|\tilde{x_i}-\tilde{x_j}\right|\textless 1}F_0\left(1-\left|\tilde{x_i}-\tilde{x_j}\right|\right)
\end{equation}
%
Through this rescaling, we have reduced the number of the parameters in our model to three: the
energy $\tilde{H}$, the interaction strength $\tilde{F_0}$, and the number of particles $N$.  The
rescaling is equivalent to setting $\sigma$, $\omega_0$ and $m$ to $1$; this is what we do in our
numerical simulation.  In the rest of this paper, we will use ``E" and ``$F_0$" to denote the
reduced versions $\tilde{H}$ and $\tilde{F_0}$.

@@

In the parameter regime where the energy $E$ is very large ($E\gg F_0$), it is dominated by the trap
energy, i.e., the interaction energy is negligible.  Thus,
$E\approx\omega_0^2\sum_{i}x_i^2=N\omega_0^2R^2$ in this regime.

\begin{figure}[tbhp]
\centering
\includegraphics[scale=0.6]{ZhiyuPictures/N=5_GS_pre_2_rev.eps}
\caption{The ground state of a system with $N=5$ particles, shown via the positions of each
  particle, labeled 1 through 5.  At large interactions, the particles position themselves just
  outside the range of interactions of the neighboring particles.}
\label{fig:GS1}
\end{figure}

@@

The lowest-energy state of the system has zero kinetic energy; in this state the particles find
stationary positions which minimize the trap (potential) and interaction energies.  The trap
potential tries to squeeze the particles towards the trap center, while the interaction tries to push
them apart.

Because of the `Heaviside theta' form of our interaction, at large enough interactions (large $F_0$)
the particles are spaced exactly at distance equal to the range $\sigma$, which is distance $1$ in
our rescaled units.  The interaction is then effectively a `hard-core' interaction.

@@

In Fig.\ref{fig:GS1}, we can see this crossover from the ``solid-like" limit (left) to the
``hardcore gas" limit (right).

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}[h]
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/stationary_and_rotating_frame_t=1-5_5T.eps}
%
\caption{Upper (lower) panels are snapshots of the cloud configuration in phase space in the
  stationary (rotating) frame, at instants within the first trap period.  Here $N=50$, $F_0=100$,
  and $E=50000$.  }
\label{fig:Breathingfrequency2_0}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.eps}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

A useful way to visualize the state and evolution of the gas is to plot the position and momentum of
each particle, i.e., to plot the locations of the particles in the single-particle phase space.
This will be useful for visualizing both the breathing mode and relaxation.

Figures \ref{fig:Breathingfrequency2_0} and \ref{fig:Breathingfrequency2_1} show such phase space
snapshots.  We generally start with particles distributed at random positions around the trap
center, initially with no velocity (zero momentum).  The initial state thus has the points lined up
along the $X$ axis.  In the absence of interactions, each particle would undergo simple harmonic
oscillation with period $T=2\pi/\omega_0 = 2\pi$.  The point corresponding to each particle than
executes clockwise elliptical motion around in the $X$-$P$ plane.  (In our units, the elliptical
trajectory is actually circular.)  This results in the initial line remaining a straight line and
rotating clockwise with exact period $T$.  The effect of interaction is to smear out the line and
spread the points out toward a rotationally invariant distribution in the  $X$-$P$ plane.  The top
panel of Figure  \ref{fig:Breathingfrequency2_0} shows this during the first period and Figure
\ref{fig:Breathingfrequency2_1} shows this process over a much longer timescale.

Since the rotation of the initial line of points is simple to understand as a single-particle
(non-interacting) effect, we can focus on interaction effects by viewing the phase space in a
`rotating' frame.  The rotating frame is used in the lower panel of Figure
\ref{fig:Breathingfrequency2_0} and in Figure \ref{fig:Breathingfrequency2_1}.  In this picture, the
`real' $X$ and $P$ axes are rotating counter-clockwise.  This rotating frame picture may be regarded
as a classical version of the ``interaction picture'' of quantum dynamics.  This picture highlights
effects of interactions because the other effects are already encoded in the frame rotation.

In the absence of interactions, each point (each particle) is stationary in the rotating-frame
phase-space picture.  As seen in lower panel of Figure \ref{fig:Breathingfrequency2_0} and in Figure
\ref{fig:Breathingfrequency2_1}, interactions cause a gradual distortion of the line as well as some
degree of rotation.
%
The rotation that is visible in the already rotating $X_{\rm rot}$-$P_{\rm rot}$ frame is the
interaction-induced shift of the breathing frequency from the noninteracting value,
$\omega=2\omega_0$.  The distortion of the line toward an eventually rotationally invariant
distribution may be regarded as thermalization or ergodicity.  In the next sections we explore these
two interaction-induced effects.

\subsection{Numerical calculations}

Since the number of particles we use in the Molecular Dynamics simulation is limited (5 to 50), the
random noise is significant. If we start with a state whose configuration in phase-space is an
ellipse slightly deviated from circle, the oscillating amplitude will be too small to be
distinguished from noise. Therefore, we start from an extreme case -- line distribution (random x,
zero p) and observe the oscillation of R(t).

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\center
\includegraphics[scale=0.32]{ZhiyuPictures/freq_scanF_differentE_log_2_with_oscillation_demo.png}
\caption{Breathing mode frequency measured at different $E$ and $F_0$. Inset: a demonstration of the
  oscillation of $E_{trap}$, the total potential energy of particles in the trap, Since
  $E_{trap}\propto \sum_{is}{x_i^2}$, it manifests the oscillation behavior of the radius $R(t)$ as
  defined in Eq.~\ref{eq:def_of_R}.  The breathing frequency is obtained from the Fourier transform
  of $R(t)$.}
\label{fig:Breathingfrequency1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

We are interested in the oscillations of the size $R(t)$ of the cloud.  It is convenient that our
finite-size simulations start with a line distribution in phase space.  If we started with a state
whose configuration in phase-space is only slightly different from a circle, the oscillating
amplitude would be too small to be distinguished from noise.

Without interaction, the breathing mode frequency is exactly 2.  This is visualized readily from the
top panel of \ref{fig:Breathingfrequency2_0}, where $R(t)$ is the extent of the distribution in the
horizontal ($X$) direction.  As the line rotates clockwise with frequency $\omega=1$, within each
period the line is twice horizontally aligned (maximum $R$) and twice vertically aligned (minimum
$R$), so that the frequency of $R(t)$ is $\omega_0=2$.  When there is interaction, the frequency
will get shifted: $\omega_0 = 2+\delta$.  We measure the radius of the cloud $R(t)$ and get the
frequency spectrum of its oscillation behavior by Fourier transform. Then we take the peak frequency
near 2 as the breathing mode frequency.

The frequency measured in this way for different $E$ and $F_0$ is shown in the
Fig.\ref{fig:Breathingfrequency1}, plotted as a function of $F_0$.  The points each correspond to
breathing-mode dynamics following from a single initial state; there is no averaging.  The curves
therefore show some randomness.  However, two prominent features are clear from these curves.  At
large interactions, $F_0\gg E$, the breathing frequency is larger than $2$, and saturates around a
value which decreases with the system energy $E$.  At small interactions, the breathing frequency is
actually smaller than the non-interacting value $2$.

Below, we will provide a detailed phase-space argument for these behaviors.  However, a simple
real-space picture explains both effects qualitatively as well.  For very large $F_0$, when the
interaction is `hard-core'-like, two particles exchange momentum immediately at a collision.  By
exchanging the labels of the particles during the collision, this can be interpreted as follows:
particle $A$ carrying momentum $P_A$ jumps by distance $\sigma$ to the right, while particle $B$
carries its momentum $P_B$ and jumps by distance $\sigma$ to the left.  In this manner, every
collision will save a particle some time, $\frac{\sigma}{v}$.  This translates into an increase of
the breathing frequency.  The speed $v$ per particle is on average $\sim\sqrt{E/N}$, and the number
of collision each particle experiences in one period of harmonic oscillation is $\sim N$; hence we
obtain $\delta\sim N^{\frac{3}{2}}E^{-\frac{1}{2}}$.  When $F_0$ is smaller, we can no longer think
of the particles as hard rods; the collisions now take finite time during which the speeds of the
particles are actually slowed down and then sped up again, as they either cross paths or bounce from
each other.  When $F_0$ is small enough, this approaching time will at some critical value of $F_0$
consume the time saved by the finite range of the interaction.  This explains why there is a
small-$F_0$ regime for which the shift $\delta$ is negative.

@@
